---
title: "Building custom themes with prefab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{prefab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Themes are functions

A prefab theme is a function that returns a `prefab_theme` object -- an ordered
list of steps. Calling a theme shows you an outline of the steps; nothing happens to the
file system until you pass the theme to `use_theme()` or
`create_project()`.

```{r inspect, eval = FALSE}
library(prefab)

# Calling r_analysis() returns a theme object -- it does not write any files
r_analysis()
#> <theme> 3 steps
#> * file → main.R (skip)
#> * file → README.md (skip)
#> * text → .gitignore (union)
#> ℹ Apply with `use_theme()` or `create_project()`
```

Because themes are functions, they can take parameters, compose with `+`, and
ship in packages.

## Applying themes

Pass a theme to `use_theme()` (existing project) or `create_project()` (new
directory):

```{r apply, eval = FALSE}
use_theme(r_analysis())
create_project("~/projects/my-targets-project", r_targets())
```

`use_theme()` discovers the project root automatically (via `.here`, `.Rproj`,
`.git`, etc.). `create_project()` creates the directory first, then applies
the theme.

## Composing themes with `+`

Themes compose with `+`, concatenating their step lists:

```{r compose, eval = FALSE}
# Project structure + Claude Code agent config
use_theme(r_analysis() + claude_r_analysis())

create_project("~/projects/new-analysis", r_targets() + claude_r_targets())
```

Order matters: later steps can override earlier ones when the merge strategy
allows it (e.g., two themes deploying the same file with `"overwrite"`).

## Writing your own theme

A theme function is any R function that returns a `prefab_theme` via
`new_theme()`. The three step types are:

- `step_file(source, dest)` -- deploy a file
- `step_text(content, dest)` -- deploy inline text
- `step_run(fn, ...)` -- execute a function for its side effects

```{r custom, eval = FALSE}
my_analysis <- function(use_data_dir = TRUE, extra_ignores = character(0)) {
  from_templates <- from_dir("~/my-templates")
  ignore_lines <- c(".Rproj.user", ".Rhistory", ".RData", extra_ignores)
  new_theme(
    from_templates("main.R", "main.R", strategy = "skip"),
    from_templates("README.md", "README.md", strategy = "skip"),
    step_text(ignore_lines, ".gitignore", strategy = "union"),
    if (use_data_dir) step_run(fs::dir_create, "data", .label = "fs::dir_create")
  )
}

use_theme(my_analysis(extra_ignores = "*.csv") + claude_r_analysis())
```

Because themes are just functions, parameters give you conditional behavior
for free. `NULL` arguments to `new_theme()` are silently dropped, so
`if (cond) step(...)` works naturally.

### Source helpers

`from_dir()` and `from_package()` create step-builders that resolve source
paths relative to a directory or an installed R package:

```{r source-helpers, eval = FALSE}
# Resolve from a local directory
from_templates <- from_dir("~/my-templates")
from_templates("header.md", "README.md", strategy = "skip")

# Resolve from an installed package's inst/ directory
from_prefab <- from_package("prefab")
from_prefab("r_analysis/main.R", "main.R", strategy = "skip")
```

`from_dir()` resolves its path to absolute at creation time, so later working
directory changes do not affect it. `from_package()` works with both installed
packages and `devtools::load_all()`.

## Distributing themes via a package

Put template files in a package's `inst/` directory and use `from_package()`:

```{r from-package, eval = FALSE}
# In mythemes/R/themes.R
#' @export
my_org_analysis <- function() {
  f <- from_package("mythemes")
  new_theme(
    f("templates/main.R", "main.R", strategy = "skip"),
    f("templates/README.md", "README.md", strategy = "skip", data = list()),
    f("claude/settings.json", ".claude/settings.json", strategy = "merge_json"),
    f("claude/rules/conventions.md", ".claude/rules/conventions.md")
  )
}
```

Users compose your themes with any others:

```{r from-package-use, eval = FALSE}
use_theme(mythemes::my_org_analysis() + prefab::claude_r_analysis())
```

## Inspecting themes with `theme_code()`

`theme_code()` prints R code that reproduces a theme -- useful for understanding
built-in themes or as a starting point for customization:

```{r theme-code, eval = FALSE}
theme_code(claude_r_analysis())
#> new_theme(
#>   from_package("prefab")("claude/settings.json", ".claude/settings.json", strategy = "merge_json"),
#>   from_package("prefab")("claude/rules/r_analysis.md", ".claude/rules/r_analysis.md"),
#>   step_text(c(".Rproj.user", ".Rhistory", ".RData", ".DS_Store"), ".gitignore", strategy = "union")
#> )
```

Copy the output, edit it into your own theme function, and swap out or add
steps. The code is also returned invisibly as a string.

## Merge strategies

Every file step declares a **strategy** for handling pre-existing destination
files. Strategies are what make `use_theme()` safe to re-run.

| Strategy | Behavior | Idempotent | Typical use |
|---|---|---|---|
| `"overwrite"` | Replace the file entirely | Yes | Managed config files |
| `"skip"` | Do nothing if file exists | Yes | Starter files users will edit |
| `"union"` | Append lines not already present | Yes | `.gitignore`, `.Rbuildignore` |
| `"append"` | Append all content unconditionally | No | Rare; prefer `"union"` |
| `"merge_json"` | Recursively merge JSON objects | Yes | `.claude/settings.json` |

Guidelines for choosing:

- Files the user should never edit: `"overwrite"`.
- Files the user will customize after first deploy: `"skip"`.
- Line-based config where entries accumulate: `"union"`.
- Structured JSON config: `"merge_json"` (objects merge key-by-key, arrays are
  union-merged, scalar collisions preserve the destination value).
- Avoid `"append"` unless you want duplicate content on re-runs.

## Template rendering

File steps support `{{var}}` interpolation when `data` is non-NULL:

```{r template, eval = FALSE}
from_templates <- from_dir("~/my-templates")
new_theme(
  # data = list() enables rendering with auto-discovered variables only
  from_templates("README.md", "README.md", strategy = "skip", data = list()),
  # Explicit variables supplement or override auto-context
  from_templates("CITATION.md", "CITATION.md",
                 data = list(org_name = "Acme Corp"))
)
```

Auto-discovered variables (built once per `use_theme()` call):

| Variable | Source |
|---|---|
| `project_dir` | Name of the project root directory |
| `package_name` | `Package` field from DESCRIPTION, or `project_dir` |
| `year` | Current year |
| `date` | Current date (`YYYY-MM-DD`) |

A template file like:

```
# {{project_dir}}

Created {{date}} by {{org_name}}.
```

is rendered by merging explicit `data` on top of auto-context. Explicit values
win on collision. If a variable is referenced but not available, rendering fails
with an informative error.

`step_text()` does not support `data` -- inline content can interpolate R
variables directly.
