---
title: "Getting started with prefab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with prefab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    if (!is.null(options$display_code)) {
      x <- options$display_code
    }
    hook_source(x, options)
  })
})
```

## Basics

The goal of prefab is to make it easier for you to set up files and directories
you need for a given project. You can think of the project scaffolding--like
README files and directory structures--as a *theme* you want to apply to a new
or existing project directory.

### Viewing and applying themes

For example, `r_analysis()` is a prefab theme shipped with this package to
provide scaffolding for a simple R data analysis project. To understand what it
does, simply call the theme to show the outline of steps:

```{r}
#| label: call_theme

library(prefab)

r_analysis()
```

Nothing happens to the file system until you pass the theme to `use_theme()` or
`create_project()`. Above, the `r_analysis()` theme indicated that it contains
five steps: two create directories, and three write files. File steps also
contain a merge strategy shown in `()` for handling pre-existing destination
files.

Pass the theme to `create_project()` to create a new project folder with that
scaffolding.

```{r, display_code='create_project(tempfile("my-analysis-project"), r_analysis())'}
#| label: create_example

withr::with_tempdir({
  app <- cli::start_app(output = "stdout", .auto_close = FALSE)
  create_project(tempfile("my-analysis-project"), r_analysis())
  cli::stop_app(app)
})
```

That created a directory called `~/my-analysis-project` and added two new
subdirectories and three new files. Then (assuming you are using Positron or
Rstudio) it opened the project in a new session.

Instead of creating a new project directory, you can also apply a theme to your
current project with `use_theme()`:

```{r, display_code="use_theme(r_analysis())"}
#| label: use_example

withr::with_tempdir({
  file.create(".here")
  app <- cli::start_app(output = "stdout", .auto_close = FALSE)
  use_theme(r_analysis())
  cli::stop_app(app)
})
```

Alternatively you may find it useful to create and launch a new project via a
keyboard shortcut mapped to `launch_project()`.

### Composing themes with `+` and arguments

A prefab theme is a function that returns an ordered list of steps that modify
files and paths, or run other functions Because themes are just functions, they
can have arguments and you can compose them with `+`, concatenating their steps.

```{r, display_code="use_theme(r_analysis(data_dirs = FALSE) + claude_r_analysis())"}
#| label: compose

withr::with_tempdir({
  file.create(".here")
  app <- cli::start_app(output = "stdout", .auto_close = FALSE)
  use_theme(r_analysis() + claude_r_analysis(settings_json = FALSE))
  cli::stop_app(app)
})
```

Steps run left to right.

## Building your own theme

There are three ways to build your own theme:

**From a directory of files.** Arrange template files in a folder and
`theme_from_dir()` turns them into a theme. For example, put your desired
template files in the folder `~/my-project-structure`:

```{r}
#| label: theme-from-dir-setup
#| echo: false

# Create a small directory of template files
theme_dir <- file.path(tempdir(), "my-project-structure")
fs::dir_create(file.path(theme_dir, "R"))
fs::dir_create(file.path(theme_dir, "data"))
writeLines(c("*.csv", ".Rhistory"), file.path(theme_dir, ".gitignore"))
writeLines("# My Project\n", file.path(theme_dir, "README.md"))
writeLines("# analysis code here", file.path(theme_dir, "R", "analysis.R"))
```

```r
├── data
├── R
│   └── analysis.R
│   .gitignore
└── README.md
```

Use `theme_from_dir()` to turn them into a theme, and then apply them to a new
or existing project:

```{r, display_code='my_theme <- theme_from_dir("~/my-project-structure")\nuse_theme(my_theme)'}
#| label: theme-from-dir-apply

my_theme <- theme_from_dir(theme_dir)

withr::with_tempdir({
  file.create(".here")
  app <- cli::start_app(output = "stdout", .auto_close = FALSE)
  use_theme(my_theme)
  cli::stop_app(app)
})
```

An optional `_prefab.yml` sidecar controls per-file merge strategies and
template data.

**By composing existing themes.** Combine and extend built-in or custom themes
with `+`:

```r
my_theme <- theme_from_dir("~/my-extras") + claude_r_analysis()
```

Themes are combined left to right.

**From steps.** Build themes programmatically with `step_file()`, `step_text()`,
and `step_run()`:

```r
my_theme <- function() {
  new_theme(
    step_file("~/my_themes/header.R", "R/header.R"),
    step_text(c("*.csv", "*.rds"), ".gitignore", strategy = "union"),
    step_run(fs::dir_create, "tables", .label = "fs::dir_create('tables')")
  )
}
```

## Writing themes from steps

A prefab theme is a function that returns an ordered list of steps that modify
files and paths, or run other functions.

The three step types are:

- `step_file(source, dest)` -- deploy a file
- `step_text(content, dest)` -- deploy inline text
- `step_run(fn, ...)` -- execute a function for its side effects

```{r}
#| label: custom
#| eval: false

my_analysis_theme <- function() {
  ignore_lines <- c(".Rproj.user", ".Rhistory", ".RData")
  new_theme(
    step_file("~/my-templates/main.R", "main.R", strategy = "skip"),
    step_text(ignore_lines, ".gitignore", strategy = "union"),
    step_run(fs::dir_create, "data", .label = "fs::dir_create")
  )
}
```

Because themes are just functions, parameters give you conditional behavior for
free. `NULL` arguments to `new_theme()` are silently dropped, so
`if (cond) step(...)` works naturally.

```{r}
#| label: custom_conditional
#| eval: false

my_analysis_theme <- function(use_data_dir = TRUE, extra_ignores = character(0)) {
  ignore_lines <- c(".Rproj.user", ".Rhistory", ".RData", extra_ignores)
  new_theme(
    step_file("~/my-templates/main.R", "main.R", strategy = "skip"),
    step_text(ignore_lines, ".gitignore", strategy = "union"),
    if (use_data_dir) step_run(fs::dir_create, "data", .label = "fs::dir_create")
  )
}
```

### Source helpers

`from_dir()` and `from_package()` create step-builders that resolve source paths
relative to a directory or an installed R package:

```{r}
#| label: source-helpers
#| eval: false

# Resolve from a local directory
from_templates <- from_dir("~/my-templates")
from_templates("header.md", "README.md", strategy = "skip")

# Resolve from an installed package's inst/ directory
from_my_package <- from_package("my_package")
from_my_package("r_analysis/main.R", "main.R", strategy = "skip")
```

`from_dir()` resolves its path to absolute at creation time, so later working
directory changes do not affect it. `from_package()` works with both installed
packages and `devtools::load_all()`.

### Merge strategies

Every file step declares a **strategy** for handling pre-existing destination
files. Strategies are what make `use_theme()` safe to re-run.

  | Strategy       | Behavior                           | Idempotent | Typical use                   |
  | -------------- | ---------------------------------- | ---------- | ----------------------------- |
  | `"overwrite"`  | Replace the file entirely          | Yes        | Managed config files          |
  | `"skip"`       | Do nothing if file exists          | Yes        | Starter files users will edit |
  | `"union"`      | Append lines not already present   | Yes        | `.gitignore`, `.Rbuildignore` |
  | `"append"`     | Append all content unconditionally | No         | Rare; prefer `"union"`        |
  | `"merge_json"` | Recursively merge JSON objects     | Yes        | `.claude/settings.json`       |

Guidelines for choosing:

- Files the user should never edit: `"overwrite"`.
- Files the user will customize after first deploy: `"skip"`.
- Line-based config where entries accumulate: `"union"`.
- Structured JSON config: `"merge_json"` (objects merge key-by-key, arrays are
  union-merged, scalar collisions preserve the destination value).
- Avoid `"append"` unless you want duplicate content on re-runs.

### Template rendering

File steps support `{{var}}` interpolation when `data` is non-NULL:

```{r}
#| label: template
#| eval: false

from_templates <- from_dir("~/my-templates")
new_theme(
  # data = list() enables rendering with auto-discovered variables only
  from_templates("README.md", "README.md", strategy = "skip", data = list()),
  # Explicit variables supplement or override auto-context
  from_templates("CITATION.md", "CITATION.md",
                 data = list(org_name = "Acme Corp"))
)
```

Auto-discovered variables (built once per `use_theme()` call):

  | Variable       | Source                                             |
  | -------------- | -------------------------------------------------- |
  | `project_dir`  | Name of the project root directory                 |
  | `package_name` | `Package` field from DESCRIPTION, or `project_dir` |
  | `year`         | Current year                                       |
  | `date`         | Current date (`YYYY-MM-DD`)                        |

A template file like:

```
# {{project_dir}}

Created {{date}} by {{org_name}}.
```

is rendered by merging explicit `data` on top of auto-context. Explicit values
win on collision. If a variable is referenced but not available, rendering fails
with an informative error.

`step_text()` does not support `data` -- inline content can interpolate R
variables directly.

## Sharing and re-using custom themes

**Source a file.** Save theme functions in an external script and source them.
This can of course be any external script, or you can put themes in
`~/.prefab-themes.R` which is sourced by the function `load_themes()`:

```r
load_themes()
use_theme(my_analysis_theme())
```

**Ship in a package.** This package ships with several basic themes, and you can
add themes (which are just functions) to your own package. Using a package is
best sharing themes across an organization.

When writing a theme function for your own package, place template files in
`inst/` and use `from_package()`.

```r
my_theme <- function() {
  from_my_package <- from_package("my_package")
  new_theme(
    from_my_package("main.R", "main.R", strategy = "skip")
    from_my_package("README.md", "README.md", strategy = "skip")
  )
}
```

## Inspecting themes with `theme_code()`

`theme_code()` prints R code that reproduces a theme -- useful for understanding
built-in themes or as a starting point for customization:

```{r}
#| label: theme-code

theme_code(claude_r_analysis())
```

Copy the output, edit it into your own theme function, and swap out or add
steps. The code is also returned invisibly as a string.
