---
title: "Getting started with prefab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with prefab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

local({
  hook_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    if (!is.null(options$display_code)) {
      x <- options$display_code
    }
    hook_source(x, options)
  })
})
```

## Basics

The goal of prefab is to make it easier for you to set up files and directories you need for a given project. You can think of the project scaffolding, like README files and directory structures, as a *theme* you want to apply to a new or existing project directory. 

For example, `r_analysis()` is a prefab theme shipped with this package to provide scaffolding for a simple R data analysis project. To understand what it does, simply call the theme to show the outline of steps:

```{r call_theme}
library(prefab)

r_analysis()
```

Nothing happens to the file system until you pass the theme to `use_theme()` or `create_project()`.
In this case, the `r_analysis()` theme shows that it contains five steps: two create directories, and three write files. File steps also contain a merge strategy shown in `()` for handling pre-existing destination files.

Passing the theme to `create_project()` will create a new project folder with that scaffolding.

```{r create_example, display_code = 'create_project(tempfile("my-analysis-project"), r_analysis())'}
withr::with_tempdir({
  app <- cli::start_app(output = "stdout", .auto_close = FALSE)
  create_project(tempfile("my-analysis-project"), r_analysis())
  cli::stop_app(app)
})
```

That created a directory called `~/my-analysis-project`, created two new subdirectories, and three new files. Then (assuming you are using Positron or Rstudio) it opened the project in a new session.

Instead of creating a new project directory, you can also apply a theme to your current project with `use_theme()`:

```{r use_example, display_code = "use_theme(r_analysis())"}
withr::with_tempdir({
  file.create(".here")
  app <- cli::start_app(output = "stdout", .auto_close = FALSE)
  use_theme(r_analysis())
  cli::stop_app(app)
})
```

## Composing themes with `+` and arguments

A prefab theme is a function that returns an ordered list of steps. In general these steps can create or modify files and paths, and also run other functions. Because themes are just functions, they can have arguments and you can compose them with `+`, concatenating their steps.

```{r compose, display_code = "use_theme(r_analysis(data_dirs = FALSE) + claude_r_analysis())"}
withr::with_tempdir({
  file.create(".here")
  app <- cli::start_app(output = "stdout", .auto_close = FALSE)
  use_theme(r_analysis() + claude_r_analysis(settings_json = FALSE))
  cli::stop_app(app)
})
```

Steps run left to right.

## Writing your own theme

There are three ways to build your own theme:

**From a directory of files.** Arrange template files in a folder and `theme_from_dir()` turns them into a theme. An optional `_prefab.yml` sidecar controls per-file merge strategies and template data.

``` r
my_theme <- theme_from_dir("~/my-extras")
```

**From steps.** Build themes programmatically with `step_file()`, `step_text()`, and `step_run()`:

``` r
my_theme <- function() {
  new_theme(
    step_file("~/my_themes/header.R", "R/header.R"),
    step_text(c("*.csv", "*.rds"), ".gitignore", strategy = "union"),
    step_run(fs::dir_create, "tables", .label = "fs::dir_create('tables')")
  )
}
```

**By composing existing themes.** Combine and extend built-in or custom themes with `+`:

``` r
my_theme <- r_analysis() + claude_r_analysis() + theme_from_dir("~/my-extras")
```

## Sharing and re-using custom themes

**Source a file.** Save theme functions in an external script and source them. `load_themes()` makes that easy: put themes in `~/.prefab-themes.R` (or set the `PREFAB_THEMES` environment variable) and call `load_themes()` to make them available.

``` r
load_themes()
use_theme(my_analysis_theme())
```

**Ship in a package.** This package ships with several basic themes, and you can add themes (which are just functions) to your own package. Use `from_package()` to resolve template files from `inst/`. Using a package is best sharing themes across an organization.

## Theme steps

A theme function is any R function that returns a `prefab_theme` via
`new_theme()`. The three step types are:

- `step_file(source, dest)` -- deploy a file
- `step_text(content, dest)` -- deploy inline text
- `step_run(fn, ...)` -- execute a function for its side effects

```{r custom, eval = FALSE}
my_analysis <- function(use_data_dir = TRUE, extra_ignores = character(0)) {
  ignore_lines <- c(".Rproj.user", ".Rhistory", ".RData", extra_ignores)
  new_theme(
    step_file("~/my-templates/main.R", "main.R", strategy = "skip"),
    from_templates("README.md", "README.md", strategy = "skip"),
    step_text(ignore_lines, ".gitignore", strategy = "union"),
    if (use_data_dir) step_run(fs::dir_create, "data", .label = "fs::dir_create")
  )
}

use_theme(my_analysis(extra_ignores = "*.csv") + claude_r_analysis())
```

Because themes are just functions, parameters give you conditional behavior
for free. `NULL` arguments to `new_theme()` are silently dropped, so
`if (cond) step(...)` works naturally.

### Source helpers

`from_dir()` and `from_package()` create step-builders that resolve source
paths relative to a directory or an installed R package:

```{r source-helpers, eval = FALSE}
# Resolve from a local directory
from_templates <- from_dir("~/my-templates")
from_templates("header.md", "README.md", strategy = "skip")

# Resolve from an installed package's inst/ directory
from_prefab <- from_package("prefab")
from_prefab("r_analysis/main.R", "main.R", strategy = "skip")
```

`from_dir()` resolves its path to absolute at creation time, so later working
directory changes do not affect it. `from_package()` works with both installed
packages and `devtools::load_all()`.

## Distributing themes via a package

Put template files in a package's `inst/` directory and use `from_package()`:

```{r from-package, eval = FALSE}
# In mythemes/R/themes.R
#' @export
my_org_analysis <- function() {
  f <- from_package("mythemes")
  new_theme(
    f("templates/main.R", "main.R", strategy = "skip"),
    f("templates/README.md", "README.md", strategy = "skip", data = list()),
    f("claude/settings.json", ".claude/settings.json", strategy = "merge_json"),
    f("claude/rules/conventions.md", ".claude/rules/conventions.md")
  )
}
```

## Inspecting themes with `theme_code()`

`theme_code()` prints R code that reproduces a theme -- useful for understanding
built-in themes or as a starting point for customization:

```{r theme-code}
theme_code(claude_r_analysis())
```

Copy the output, edit it into your own theme function, and swap out or add
steps. The code is also returned invisibly as a string.

## Merge strategies

Every file step declares a **strategy** for handling pre-existing destination
files. Strategies are what make `use_theme()` safe to re-run.

| Strategy | Behavior | Idempotent | Typical use |
|---|---|---|---|
| `"overwrite"` | Replace the file entirely | Yes | Managed config files |
| `"skip"` | Do nothing if file exists | Yes | Starter files users will edit |
| `"union"` | Append lines not already present | Yes | `.gitignore`, `.Rbuildignore` |
| `"append"` | Append all content unconditionally | No | Rare; prefer `"union"` |
| `"merge_json"` | Recursively merge JSON objects | Yes | `.claude/settings.json` |

Guidelines for choosing:

- Files the user should never edit: `"overwrite"`.
- Files the user will customize after first deploy: `"skip"`.
- Line-based config where entries accumulate: `"union"`.
- Structured JSON config: `"merge_json"` (objects merge key-by-key, arrays are
  union-merged, scalar collisions preserve the destination value).
- Avoid `"append"` unless you want duplicate content on re-runs.

## Template rendering

File steps support `{{var}}` interpolation when `data` is non-NULL:

```{r template, eval = FALSE}
from_templates <- from_dir("~/my-templates")
new_theme(
  # data = list() enables rendering with auto-discovered variables only
  from_templates("README.md", "README.md", strategy = "skip", data = list()),
  # Explicit variables supplement or override auto-context
  from_templates("CITATION.md", "CITATION.md",
                 data = list(org_name = "Acme Corp"))
)
```

Auto-discovered variables (built once per `use_theme()` call):

| Variable | Source |
|---|---|
| `project_dir` | Name of the project root directory |
| `package_name` | `Package` field from DESCRIPTION, or `project_dir` |
| `year` | Current year |
| `date` | Current date (`YYYY-MM-DD`) |

A template file like:

```
# {{project_dir}}

Created {{date}} by {{org_name}}.
```

is rendered by merging explicit `data` on top of auto-context. Explicit values
win on collision. If a variable is referenced but not available, rendering fails
with an informative error.

`step_text()` does not support `data` -- inline content can interpolate R
variables directly.


